---
title: "Spatial Transcriptomics Analysis: Human Brain"
author: "Vishishtaa Pandit"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

## 1. Introduction

Spatial transcriptomics preserves the physical location of gene 
expression within intact tissue sections. Unlike bulk RNA-seq, 
which loses spatial information by homogenising tissue, spatial 
transcriptomics allows us to ask not just *what* genes are 
expressed, but *where* in the tissue they are active.

This analysis uses the 10X Genomics Visium platform data from 
human brain tissue (V1_Human_Brain_Section_1) to:

1. Identify transcriptionally distinct spatial domains
2. Map known cell type markers to anatomical regions
3. Calculate spatial senescence burden across brain regions
4. Cross-reference muscle ageing DEGs with brain spatial patterns

## 2. Setup

```{r libraries}
library(Seurat)
library(tidyverse)
library(ggplot2)
library(patchwork)
```

## 3. Load Data

```{r load_data}
# Load 10X Visium data
brain <- Load10X_Spatial(
  data.dir = "data/",
  filename = "V1_Human_Brain_Section_1_filtered_feature_bc_matrix.h5"
)

cat("Dataset dimensions:\n")
cat("Genes:", nrow(brain), "\n")
cat("Spots:", ncol(brain), "\n")
```

## 4. Quality Control

```{r qc_plots, fig.width=12}
# Calculate mitochondrial percentage
brain[["percent.mt"]] <- PercentageFeatureSet(
  brain, pattern = "^MT-"
)

# Visualise QC metrics on tissue
p_counts <- SpatialFeaturePlot(
  brain, 
  features = "nCount_Spatial",
  pt.size.factor = 1.2
) + ggtitle("Total UMI Counts per Spot")

p_features <- SpatialFeaturePlot(
  brain,
  features = "nFeature_Spatial", 
  pt.size.factor = 1.2
) + ggtitle("Genes Detected per Spot")

p_counts + p_features
```

```{r save_qc, include=FALSE}
png("figures/00_qc_spatial.png",
    width = 14, height = 6, units = "in", res = 300)
print(p_counts + p_features)
dev.off()
```

```{r filter}
# Filter low quality spots
brain <- subset(
  brain,
  subset = nCount_Spatial > 500 &
           nFeature_Spatial > 200 &
           percent.mt < 25
)

cat("Spots after QC filtering:", ncol(brain), "\n")
```

## 5. Normalisation

```{r normalise}
# SCTransform normalisation — accounts for 
# sequencing depth differences between spots
brain <- SCTransform(
  brain,
  assay = "Spatial",
  verbose = FALSE
)
```

SCTransform uses a regularised negative binomial regression 
to normalise gene expression, accounting for technical 
variation between spots while preserving biological signal.

## 6. Dimensionality Reduction and Clustering

```{r clustering}
brain <- RunPCA(brain, assay = "SCT", verbose = FALSE)
brain <- FindNeighbors(brain, dims = 1:30, verbose = FALSE)
brain <- FindClusters(brain, resolution = 0.5, verbose = FALSE)
brain <- RunUMAP(brain, dims = 1:30, verbose = FALSE)

cat("Number of clusters found:", 
    length(unique(brain@meta.data$seurat_clusters)), "\n")
```

```{r cluster_plots, fig.width=14}
p_umap <- DimPlot(
  brain, 
  reduction = "umap", 
  label = TRUE,
  label.size = 5
) + 
ggtitle("UMAP — Transcriptional Clusters") +
theme_minimal()

p_spatial <- SpatialDimPlot(
  brain, 
  label = TRUE, 
  label.size = 3
) + 
ggtitle("Spatial Cluster Distribution")

p_umap + p_spatial
```

```{r save_clusters, include=FALSE}
png("figures/01_umap_spatial_clusters.png",
    width = 14, height = 6, units = "in", res = 300)
print(p_umap + p_spatial)
dev.off()
```

The UMAP (left) shows transcriptional relationships between 
spots regardless of location. The spatial plot (right) maps 
these clusters onto the actual tissue, revealing that 
transcriptionally similar spots occupy anatomically coherent 
regions — validating the biological relevance of our clustering.

## 7. Cluster Marker Genes

```{r find_markers}
markers <- FindAllMarkers(
  brain,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)

markers$cluster <- as.factor(markers$cluster)

top_markers <- do.call(rbind, lapply(
  levels(markers$cluster), function(cl) {
    sub <- markers[markers$cluster == cl, ]
    sub <- sub[order(-sub$avg_log2FC), ]
    head(sub, 5)
  }
))

print(top_markers)
```

```{r marker_interpretation}
# Known brain cell type markers
cat("=== Cluster Biological Interpretation ===\n\n")
cat("Cluster 0 — MBP, MOBP present:\n")
cat("  → Likely WHITE MATTER / Oligodendrocytes\n")
cat("  MBP (Myelin Basic Protein) is a hallmark of myelinating cells\n\n")
cat("Cluster 1 — CARTPT, CALB1 present:\n") 
cat("  → Likely specific NEURONAL populations\n")
cat("  CALB1 (Calbindin) marks inhibitory interneurons\n\n")
```

```{r marker_spatial, fig.width=12, fig.height=10}
# Visualise key markers spatially
p_markers <- SpatialFeaturePlot(
  brain,
  features = c("MBP", "MOBP", "CARTPT", "CALB1"),
  ncol = 2,
  pt.size.factor = 1.2
)

print(p_markers)
```

```{r save_markers, include=FALSE}
ggsave("figures/02_marker_genes_spatial.png",
       plot = p_markers,
       width = 12, height = 10,
       dpi = 300)

write.csv(markers,
  "results/cluster_markers.csv", row.names = FALSE)
write.csv(top_markers,
  "results/top5_markers_per_cluster.csv", row.names = FALSE)
```

## 8. Spatial Senescence Burden

Cellular senescence — where cells permanently exit the cell 
cycle — is a hallmark of ageing. Here we calculate a 
per-spot senescence score using a curated gene signature 
and map it spatially to identify brain regions with highest 
senescence burden.

```{r senescence_score}
# Curated senescence gene signature (SenMayo/CellAge)
# Curated senescence genes
senescence_genes <- c(
  "CDKN1A", "CDKN2A", "TP53", "MDM2",
  "IL6", "CXCL1", "MMP3", "SERPINE1",
  "IGFBP3", "IGFBP5", "GLB1", "LMNB1",
  "CCND1", "RB1", "E2F1", "PCNA",
  "BCL2", "BAX", "CASP3", "ATM"
)

# Filter to genes present
sen_present <- senescence_genes[
  senescence_genes %in% rownames(brain)
]

cat("Senescence genes found:", length(sen_present), "\n")
cat("Genes:", paste(sen_present, collapse = ", "), "\n")

# Only run if enough genes found
if (length(sen_present) >= 5) {
  brain <- AddModuleScore(
    brain,
    features = list(sen_present),
    name = "Senescence_Score",
    nbin = 10  # reduce bins for small gene sets
  )
  cat("Senescence score calculated successfully\n")
} else {
  cat("Too few senescence genes found — skipping module score\n")
  brain$Senescence_Score1 <- 0
}
```

```{r senescence_plot, fig.width=12}
p_sen <- SpatialFeaturePlot(
  brain,
  features = "Senescence_Score1",
  pt.size.factor = 1.3
) +
scale_fill_gradient2(
  low = "#2196F3",
  mid = "white",
  high = "#F44336",
  midpoint = 0
) +
ggtitle("Spatial Senescence Burden — Human Brain") +
theme(legend.position = "right")

print(p_sen)
```

```{r save_senescence, include=FALSE}
ggsave("figures/03_senescence_score_spatial.png",
       plot = p_sen,
       width = 10, height = 8,
       dpi = 300)
```

## 9. Cross-Tissue Ageing Analysis

Here we ask a novel question: do genes upregulated with 
ageing in skeletal muscle (identified in our companion 
bulk RNA-seq analysis) show spatially restricted expression 
patterns in brain tissue?

```{r cross_tissue}
# Load muscle ageing DEGs from companion analysis
muscle_file <- "../deseq2-aging-analysis/results/significant_DEGs.csv"

if (file.exists(muscle_file)) {
  muscle_degs <- read.csv(muscle_file)
  
  library(org.Hs.eg.db)
  
  # Convert ENSEMBL to gene symbols
  muscle_symbols <- mapIds(
    org.Hs.eg.db,
    keys = muscle_degs$gene,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
  
  # Find overlap with brain spatial data
  muscle_in_brain <- na.omit(muscle_symbols[
    muscle_symbols %in% rownames(brain)
  ])
  
  cat("Muscle ageing DEGs:", nrow(muscle_degs), "\n")
  cat("Found in brain spatial data:", 
      length(muscle_in_brain), "\n")
  
  # Split by direction
  up_ensembl <- muscle_degs %>%
    filter(log2FoldChange > 0) %>%
    pull(gene)
  
  up_symbols <- na.omit(muscle_symbols[up_ensembl])
  up_in_brain <- up_symbols[up_symbols %in% rownames(brain)]
  
  down_ensembl <- muscle_degs %>%
    filter(log2FoldChange < 0) %>%
    pull(gene)
  
  down_symbols <- na.omit(muscle_symbols[down_ensembl])
  down_in_brain <- down_symbols[down_symbols %in% rownames(brain)]
  
  cat("Upregulated in aged muscle, found in brain:", 
      length(up_in_brain), "\n")
  cat("Downregulated in aged muscle, found in brain:", 
      length(down_in_brain), "\n")
  
  # Score brain spots
  if (length(up_in_brain) > 0) {
    brain <- AddModuleScore(
      brain,
      features = list(up_in_brain),
      name = "MuscleAging_Up",
      nbin = 10
    )
  }
  
  if (length(down_in_brain) > 0) {
    brain <- AddModuleScore(
      brain,
      features = list(down_in_brain),
      name = "MuscleAging_Down",
      nbin = 10
    )
  }
  
} else {
  cat("Muscle DEG file not found.\n")
  cat("Run the DESeq2 aging analysis first.\n")
}
```

```{r cross_tissue_plot, fig.width=14}
if (file.exists(muscle_file) & length(up_in_brain) > 0) {
  
  p_up <- SpatialFeaturePlot(
    brain,
    features = "MuscleAging_Up1",
    pt.size.factor = 1.2
  ) +
  scale_fill_gradient2(
    low = "#2196F3", mid = "white", high = "#F44336",
    midpoint = 0
  ) +
  ggtitle("Muscle Ageing Upregulated Genes in Brain")
  
  p_down <- SpatialFeaturePlot(
    brain,
    features = "MuscleAging_Down1",
    pt.size.factor = 1.2
  ) +
  scale_fill_gradient2(
    low = "#F44336", mid = "white", high = "#2196F3",
    midpoint = 0
  ) +
  ggtitle("Muscle Ageing Downregulated Genes in Brain")
  
  print(p_up + p_down)
  
  ggsave("figures/04_cross_tissue_aging.png",
         plot = p_up + p_down,
         width = 14, height = 6,
         dpi = 300)
}
```

## 10. Spatially Variable Genes

```{r svg, fig.width=14, fig.height=10}
# Find spatially variable genes using Moran's I
brain <- FindSpatiallyVariableFeatures(
  brain,
  assay = "SCT",
  features = VariableFeatures(brain)[1:1000],
  selection.method = "moransi"
)

top_svg <- SpatiallyVariableFeatures(
  brain,
  method = "moransi"
)[1:6]

cat("Top spatially variable genes:\n")
print(top_svg)

p_svg <- SpatialFeaturePlot(
  brain,
  features = top_svg,
  ncol = 3,
  pt.size.factor = 1.2
)

print(p_svg)
```

```{r save_svg, include=FALSE}
ggsave("figures/05_spatially_variable_genes.png",
       plot = p_svg,
       width = 14, height = 10,
       dpi = 300)
```

## 11. Summary

```{r summary}
cat("=== Analysis Summary ===\n")
cat("Dataset: V1 Human Brain Section 1 (10X Visium)\n")
cat("Spots after QC:", ncol(brain), "\n")
cat("Genes analysed:", nrow(brain), "\n")
cat("Clusters identified:", 
    length(unique(brain@meta.data$seurat_clusters)), "\n")
cat("Senescence genes mapped:", length(sen_present), "\n")

if (exists("muscle_in_brain")) {
  cat("Muscle ageing DEGs found in brain:", 
      length(muscle_in_brain), "\n")
}

cat("\nKey finding:\n")
cat("Transcriptional clusters correspond to anatomically\n")
cat("coherent brain regions, with white matter clearly\n") 
cat("delineated by MBP/MOBP expression.\n")
```